\documentclass[a4paper,english]{article}
\usepackage[pdftex]{graphicx}
\setlength{\parskip}{\medskipamount}
\setlength{\parindent}{0pt}

% Uncomment the following line to allow the usage of umlauts and other non-ASCII characters
%\usepackage[utf8]{inputenc}

\begin{document}

\title{The CINF UHV companion}

\maketitle

\section{Vacuum}
In the high vacuum world, pressures are normally measured in the rather arbitrary units mbar and torr, which only differ by a factor of 1.33 which is a number close enough to 1 that the two units are used more or less interchangeable. A normal UHV system will typically have a base pressure on the order of $10^{-11}$ to $10^{-10}$\,mbar. Using rather simple and reasonable assumptions it can be shown that at a pressure of $10^{-6}$\,mbar, every surface atom in the vacuum chamber will on average be hit by a molecule once every second. As a rule of thumb this means that at $10^{-10}$ you can expect a clean sample to stay clean for approximately one hour.

\section{Flanges}
Every vacuum chamber is surrounded by air. The pressure of ambient air is $10^3$\,mbar or some 14 orders of magnitude higher than inside the chamber. This means that air is trying significantly hard to get inside the chamber, and one must show quite a bit of carefulness to avoid this. Normally every flange is packed with a copper gasket which is deformed from both sides by knifes mounted on the flanges. It is this deformation that ensures that the copper will make a perfect seal to both flanges. In order to achieve this, the bolts should be tightened reasonably hard, but not too much. Tightening too hard will only mean, that it is hard to open the flanges and that you loose the possibility to tighten a bit extra in  case of a leak. You should be careful to not tighten one bolt at a time, but distribute the applied force between the bolts to make sure you do not deform the gasket.

\section{The content of air}
In daily life, air might seem to be a harmless substance, but in reality it is terribly dirty, and you certainly do not want to have it inside your chamber. At 1\,bar every molecule that exists in concentrations of just 1\,ppb will hit every surface atom once every second, basically meaning, that if you chamber has seen air, you should regard it as polluted with basically every organic compound know to man as well as water and oxygen. Even at very small air-pressures, as can be seen through a leak, atleast water and oxygen will be available in sufficient amounts to be unacceptable doing experiments.

\section{Leak testing}
Leak testing is the art of finding the origin of a leak in the chamber. The way to attack this problem depends heavily on the size of the suspected leak. If the chamber is reasonably tight so that it can be pumped to below $10^{-5}$\,mbar, one can turn on the mass-spectrometer. In this case leak-testing with helium is the preferred method. Helium is more or less the perfect molecule for this task since it a) is only existing in absolutely minute amounts in air, b) it is the smallest molecule in Nature and therefore the one that will most easily slip through a leak. The procedure is then to measure mass 4 on the mass-spec while blowing helium on the chamber one flange at a time, if you get a signal on the mass-spec, you have found your leak. Remember that helium rises in air and for that reason, it is wise to start from the top of the chamber. If the chamber does not have a mass-spectrometer, the ion gauge can be used to find at least rather large leaks. This works because the ionisation energy of helium is much larger than that of both oxygen and nitrogen, and  therefore you will see a drop in the signal from the ion gauge, when hellium out-competes air in the vicinity of the leak.

If the leak is large enough that the chamber cannot be pumed low enough to turn on the mass-spec, life is a bit more difficult. In that case one is more or less forced to randomly tighten flanges under suspicion of being leak. If this also fails, another option is to back-fill the chamber with a rather large amount of hellium and use an outside sniffer connected to the leak-tester to find the leak.

Once the leak is found, often the problem can be solved simply by tightening the flange. If this does not solve the problem, chances are, that the gasket is not mounted properly and you will have to open the flange.  

\section{UHV pumps}

\subsection{Turbo pumps}
The turbo pump is the true workhorse in a UHV-system. It works by spinning plates tilted at an angle very fast and thus it mechanically kicks the molecules out of the chamber, once the molecule is out of the pump, it reaches a rouging pump, that takes it the rest of the way to atmospheric pressure. For this kind of pum to work, the pressure needs to be low enough that the molecule-molecule interactions are very small, meaning somewhere below $10^{0}$\,mbar. A turbo pump is a more or less perfect pump in the sense that practically no back-flow occurs --- no gas from the pre-vacuum side will enter the chamber. This gives the very nice property that even though it will of course not provide a perfect vacuum, the turbo pump will provide a very clean vacuum consisting only of the gasses that are already in chamber. Thus it is possible to have a residual gas mixture containing only a few well known species.

A turbo pump is not made to move large amount of gas, but it will work reliably up to at least $10^{-2}$\,mbar, even for extended amounts of time, provided sufficient cooling and a good roughing pump capable of removing the gas from the backside of the pump fast enough. The workload of the pump can to some degree be measured from the current drawn by the pump, which can normally be seen on the control unit. A turbo pump should typically not be using more than 1-1,2\,A for extended amounts of time.

On a practical note, a turbo pump is spinning very fast (typically between 800 and 1500\,Hz depending on the size of the pump) which means that you shuld NOT try to move a spinnig turbo unless you want to perform a very expensive demonstration of applied mechanics. Most pumps are by nature stationary but for instance the leak-tester has a turbo-pump that can in principle be moved.

\subsection{Ion pumps}
An ion pump is closed pump, meaning that it will store the pumped molecules inside the pump itself. This obviously means that ion pumps should only be used at rather low pressures, typically below $10^{-7}$\,mbar --- at such low pressures it is possible to accumulate the pumped gasses even for extended amounts of time. Once in a while (typically at unpredictable times) the ion pump will release the enclosed gass, which must then be pumped away with a turbo pump. Ion pumps provides excellent base pressures and also very convienintly have no moving parts.

\subsection{TSPs}
A TSP is also a closed type of pump. It works by evaporating titanium on the surface of the pump itself. This titanium will then very efficiently bind some of the residual gass in the chamber. Next time the pump is run, most of this gass will be released but in much more concentrated form, meaning that it will be pumped more efficiently by the other pumps in the system. A TSP is very good at 'cleaning up' after an experiment that left the camber more dirty than what is normally wanted.

\section{Roughing pumps}
A roughing pumps purpose in life when placed in a UHV lab is normally to act as backing for the turbo pump. At CINF you will find typically two main kinds, the oil-lubricated pumps and the non-lubricated pumps (so-called Scroll pumps). The oil-lubricated kind is by far the most common one. It will normally provide a vacuum between $10^{-2}$ and $10^{-3}$\,mbar, preferably closer to the latter value. If the pump is not able to reach $5\times10^{-2}$ it is most likely in need of service. A scroll pump will not reach nearly the same ultimate pressure as a oil-lubricated pump. Normally one should keep an eye on the pressure of the roughing pump, it should stay quite constant, and any sign of an increased pressure is properly a sign of a pump that should be serviced.

An extremely important thing to be aware of, is the composition of the residual gas in a roughing pump. For an oil-lubricated pump, the residual gas consist mostly of various nasty organic compounds and for a scroll pump it consists mostly of air(?). In both cases the residual gas must be considered to be extremely dirty, why it is important that you never apply the pump in a way that will allow for backflow from the pump to reach your nice and clean vacuum chamber. In normal operation this is ensured by the turbo pump, but in situations where the roughing pump is pumping directly on the chamber, one needs to think. If you are pumping down a vented chamber, everything is ok, you are going to bake the chamber anyway. If you are pumping down a system considered clean (high-pressure cells, load-locks and gas-lines come to mind) you must NOT pump too low in pressure with the roughing pump. Stay in a pressure range no lower than 2 to 10\,mbar, which will ensure a high enough flow into the pump, that essentially no residual gas will back-flow. From this pressure you can leak the gas into a turbo pump using a good quality valve.

\section{Valves}
At CINF you will typically find a certain variaty of valves. Three main kinds exists as well as a number of other less common types. The three main types are the VAT valve, the Swagelok valve and the ???-valve. The swagelock and to a certain extent the VAT valve has the rather nice property, that the can be opened in a controlled way, allowing for a slow opening between two compartments. This is very practical when you want to leak a semi-high pressure into a turbo pump, which is for instance the case after pumping down a clean system with a roughing pump.
Everytime one operates a valve, one needs to be aware of the current conditions on the two sides of the valve. Often one side of the valve is exposed to the vacuum and the other side may or may not also be pumped down. If it is not, and you open the valve, you will very quickly get in to trouble, the gass from the high pressure side will rush into the vacuum part and an infinite amount of problems will arise.

\section{Baking}
After a chamber has been vented, it has as mentioned been exposed to basically any unpleasant gas, one can imagine. All this pollution needs to be removed. Some substances is easily removed just by pumping, but others forms layers on the inside of chamber, which is not easily pumped away. Water and pumping oil are two notable substances showing this behaviour. Even though the gasses will of course eventually be pumped away, the time constant for this is basically infinite and something must be done to increase the speed. The solution is to increase the temperature of the entire chamber, so-called baking. Normally baking is perform at temperatures around 150C, at these temperatures, the vapour pressure of the contaminants rises significantly, and thus it will desorp away from the walls an into the volume of the chamber, where it can be pumped efficiently. After a few days of baking, your chamber should be reasonable clean.
An important thing to remember when baking is that you should always bake all parts of a given volume at the same time. If any part of the chamber is not heated, this part will act as a cold finger where all the desorbed contaminants will really like to re-adsorb. Once the chamber is cooled, the cold spot will soon equilibrate with the rest of the chamber, and your baking is ruined.

% Uncomment the following two lines if you want to have a bibliography. Please do not forget to add an entry to your bibliography and reference it by using the \cite{} command
%\bibliographystyle{alphadin}
%\bibliography{document}

% End of the document
\end{document}

